### 2015-12-07 paper/3-features.tex
Typo.

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
We now discuss a number of bottlenecks that were discovered during the profiling of several benchmarks.
\subsubsection{Heap}
In dxex, events are always passed with pointers and thus involve a possibly expensive copy of the payload caused by allocation and copying overhead. In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}. In this way, allocating threads do not block each other. If desired, arena-pools are available for single-threaded simulation. A disadvantage is increased complexity in ownership semantics. The creating kernel is responsible for destruction, but this can only be guaranteed up to the GVT. Experiments with synchronized pools proved to be slower than implementations with the standard malloc/free.\\
Initially, dxes was using strings as identifiers as was also done in PyPDEVS. Profiling quickly indicated that this caused a real performance bottleneck. In C++, strings are heap allocated variable sized objects with an atomic reference count and not not immutable objects as in Python. Access of that reference count across threads turned out to be quite expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects or copy existing objects.\\
Strings, however, are more intuitive to work with from a user's perspective, so, as a compromise, we allowed the user to reference models/ports by string name. Once the simulation starts however, all objects use integral identifiers. This also increased the usage of the constexpr feature of C++11 in, amongst others, time stamps and message headers.
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting became very expensive, and the (de)allocation of memory caused significant contention between threads. So for the models and the kernels, dxex is still using smart pointers whereas for the messeages, a raw pointer to compacted memory is being used. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations and, occasionally, we can leverage memory orderings to only pay for synchronization when we need it. Messages, on the other hand, are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the time stamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers to choose from with varying performance depending on the simulation type. Profiling showed that, the heap implementation used in adevs was faster than any of the schedulers we had tested before (in a C++ environment). Unlike most node based heaps, this scheduler uses a fixed size array where its heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.

>>>>

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
We now discuss a number of bottlenecks that were discovered during the profiling of several benchmarks.
\subsubsection{Heap}
In dxex, events are always passed with pointers and thus avoid a possibly expensive copy of the payload caused by allocation and copying overhead. In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}. In this way, allocating threads do not block each other. If desired, arena-pools are available for single-threaded simulation. A disadvantage is increased complexity in ownership semantics. The creating kernel is responsible for destruction, but this can only be guaranteed up to the GVT. Experiments with synchronized pools proved to be slower than implementations with the standard malloc/free.\\
Initially, dxes was using strings as identifiers as was also done in PyPDEVS. Profiling quickly indicated that this caused a real performance bottleneck. In C++, strings are heap allocated variable sized objects with an atomic reference count and not not immutable objects as in Python. Access of that reference count across threads turned out to be quite expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects or copy existing objects.\\
Strings, however, are more intuitive to work with from a user's perspective, so, as a compromise, we allowed the user to reference models/ports by string name. Once the simulation starts however, all objects use integral identifiers. This also increased the usage of the constexpr feature of C++11 in, amongst others, time stamps and message headers.
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting became very expensive, and the (de)allocation of memory caused significant contention between threads. So for the models and the kernels, dxex is still using smart pointers whereas for the messeages, a raw pointer to compacted memory is being used. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations and, occasionally, we can leverage memory orderings to only pay for synchronization when we need it. Messages, on the other hand, are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the time stamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers to choose from with varying performance depending on the simulation type. Profiling showed that, the heap implementation used in adevs was faster than any of the schedulers we had tested before (in a C++ environment). Unlike most node based heaps, this scheduler uses a fixed size array where its heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.
